Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-02 _tex/index.tex
Built site for gh-pages

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021, Bouwer_et_al_2023},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the mechanisms generating the DCJ
outcome. This study aims to fill this gap by representing DCJ within the
framework of causal inference. Specifically, utilizing the structural
approach \citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study
develops a scientific model to clarify plausible causal assumptions and
mechanisms inherent in the DCJ system. The study then translates the
scientific model into a probabilistic statistical model. This model aims
to produce statistical estimates to draw inferences about plausible
causal relationships within the DCJ system.

@@

Several approaches to causal inference and counterfactuals exist, but
two are particularly prominent: the potential outcomes approach, also
known as the Neyman-Rubin causal model
\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
\citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}. Both approaches
employ rigorous mathematical notation to characterize causal inference,
but they do so in different ways \citep{Neal_2020}. The potential
outcomes approach relies on counterfactual notation, whereas the
structural approach employs the do-operator and structural causal models
\citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. Despite these differences,
both notations can be expressed in terms of the other, and both
approaches provide methods for using experimental and observational data
to estimate causal effects \citep{Pearl_2010}.

@@

One key advantage of DAGs is that, regardless of complexity, they can
represent various causal structures using only five fundamental building
blocks \citep{Neal_2020, McElreath_2020}. Figure~\ref{fig-dags}
illustrates these blocks. Figure~\ref{fig-dag_bb1} depicts two
unconnected nodes, representing an scenario where variables \(X_{1}\)
and \(X_{2}\) are not causally related. Figure~\ref{fig-dag_bb2} shows
two connected nodes, illustrating a scenario where a parent node
\(X_{1}\) exerts a causal influence on a child node \(X_{2}\).
Consequently, \(X_{2}\) is considered a \emph{descendant} of \(X_{1}\).
Figure~\ref{fig-dag_bb3} depicts a \emph{chain} (or \emph{pipe}), where
\(X_{1}\) influences \(X_{2}\), and \(X_{2}\) influences \(X_{3}\). In
this configuration, \(X_{1}\) is a parent node of \(X_{2}\), and
\(X_{2}\) a parent node of \(X_{3}\). Furthermore, the DAG shows that
\(X_{1}\) is an \emph{ancestor} of \(X_{3}\) and that the relationship
between these variables is entirely \emph{mediated} by \(X_{2}\).
Figure~\ref{fig-dag_bb4} illustrates a \emph{fork}, where variables
\(X_{1}\) and \(X_{3}\) are both influenced by \(X_{2}\). In this
scenario \(X_{2}\) is a parent node of both \(X_{1}\) and \(X_{3}\).
Finally, Figure~\ref{fig-dag_bb5} depicts a \emph{collider}, also known
as \emph{inmorality}, where variables \(X_{1}\) and \(X_{3}\) are
concurrent causes of \(X_{2}\). In this configuration, \(X_{1}\) and
\(X_{3}\) are not causally related to each other but both influence
\(X_{2}\).

@@

\includegraphics[width=0.85\textwidth,height=\textheight]{images/figures/dag_bb3.png}

@@

\includegraphics[width=0.85\textwidth,height=\textheight]{images/figures/dag_bb4.png}

@@

\includegraphics[width=0.85\textwidth,height=\textheight]{images/figures/dag_bb5.png}

@@

Given the heuristic nature of DAGs, a motivating example can help
illustrate the use of the five fundamental building blocks to construct
a system's causal structure. This example can also help to clarify
additional conventions for drawing DAGs.

@@

Figure~\ref{fig-example} presents two graphs illustrating the plausible
causal structure of the motivating example.
Figure~\ref{fig-dag_example1} shows the \emph{simplified} DAG. In this
graph, the variables of interest or \emph{endogenous} variables
\(V=\{T,X,Y\}\) are depicted as solid black circles, indicating that
they are observed variables. The arrows in the graph reflect the
expected direction of causal influences among the variables.
Figure~\ref{fig-dag_example2}, in contrast, depicts the \emph{magnified}
version of the DAG. This graph includes, in addition to the endogenous
variables, the \emph{exogenous} variables \(E=\{e_{T},e_{X},e_{Y}\}\).
These exogenous variables are commonly referred to as
\emph{disturbances} or \emph{errors}, because they represent factors not
explicitly modeled. They are depicted as open circles, indicating their
unobserved nature. While such exogenous variables are often omitted for
simplicity, including them in the graph can be advantageous in certain
scenarios, as they can help to highlight potential issues related to
conditioning and confounding \citep{Cinelli_et_al_2020}, concepts that
will be discussed in the next section.

@@

Furthermore, a detailed examination of both figures reveals the presence
of at least four of the five fundamental building blocks.
Figure~\ref{fig-dag_example1} displays multiple descendants, evident in
pairwise relations such as \(X \rightarrow T\), \(X \rightarrow Y\), and
\(T \rightarrow Y\). It also illustrates a fork with
\(X \rightarrow \{T,Y\}\). Similarly, Figure~\ref{fig-dag_example2}
features multiple two-unconnected nodes, evident in the pairwise
relations \(\{e_{T} \perp\!\!\!\perp e_{X}\}\),
\(\{e_{T} \perp\!\!\!\perp e_{Y}\}\), and
\(\{e_{X} \perp\!\!\!\perp e_{Y}\}\), as well as colliders such as
\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\). The
symbol \(\perp\!\!\!\perp\) denotes a concept know as
\emph{d-separation}, which roughly implies the independence of the
variables. This concept will be introduced in the next section.

>>>>

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021, Bouwer_et_al_2023},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the mechanisms generating the DCJ
outcome. This study aims to fill this gap by representing DCJ within the
framework of causal inference. Specifically, utilizing the structural
approach, the study develops a scientific model to clarify plausible
causal assumptions and mechanisms inherent in the DCJ system. The study
then translates the scientific model into a probabilistic statistical
model. This model aims to produce statistical estimates to draw
inferences about plausible causal relationships within the DCJ system.

@@

Several approaches to causal inference and counterfactuals exist, but
two are particularly prominent: the potential outcomes approach, also
known as the Neyman-Rubin causal model
\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
\citep{Pearl_2009, Pearl_et_al_2016}. Both approaches employ rigorous
mathematical notation to characterize causal inference, but they do so
in different ways \citep{Neal_2020}. The potential outcomes approach
relies on counterfactual notation, whereas the structural approach
employs the do-operator and structural causal models
\citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. Despite these differences,
both notations can be expressed in terms of the other, and both
approaches provide methods for using experimental and observational data
to estimate causal effects \citep{Pearl_2010}.

@@

One key advantage of DAGs is that, regardless of complexity, they can
represent various causal structures using only five fundamental building
blocks \citep{Neal_2020, McElreath_2020, Morgan_et_al_2014}.
Figure~\ref{fig-dags} illustrates these blocks. Figure~\ref{fig-dag_bb1}
depicts two unconnected nodes, representing an scenario where variables
\(X_{1}\) and \(X_{2}\) are not causally related.
Figure~\ref{fig-dag_bb2} shows two connected nodes, illustrating a
scenario where a parent node \(X_{1}\) exerts a causal influence on a
child node \(X_{2}\). Consequently, \(X_{2}\) is considered a
\emph{descendant} of \(X_{1}\). Figure~\ref{fig-dag_bb3} depicts a
\emph{chain} (or \emph{pipe}), where \(X_{1}\) influences \(X_{2}\), and
\(X_{2}\) influences \(X_{3}\). In this configuration, \(X_{1}\) is a
parent node of \(X_{2}\), and \(X_{2}\) a parent node of \(X_{3}\).
Furthermore, the DAG shows that \(X_{1}\) is an \emph{ancestor} of
\(X_{3}\) and that the relationship between these variables is entirely
\emph{mediated} by \(X_{2}\). Figure~\ref{fig-dag_bb4} illustrates a
\emph{fork}, where variables \(X_{1}\) and \(X_{3}\) are both influenced
by \(X_{2}\). In this scenario \(X_{2}\) is a parent node of both
\(X_{1}\) and \(X_{3}\). Finally, Figure~\ref{fig-dag_bb5} depicts a
\emph{collider}, also known as \emph{inmorality}, where variables
\(X_{1}\) and \(X_{3}\) are concurrent causes of \(X_{2}\). In this
configuration, \(X_{1}\) and \(X_{3}\) are not causally related to each
other but both influence \(X_{2}\).

@@

\includegraphics[width=0.75\textwidth,height=\textheight]{images/figures/dag_bb3.png}

@@

\includegraphics[width=0.75\textwidth,height=\textheight]{images/figures/dag_bb4.png}

@@

\includegraphics[width=0.75\textwidth,height=\textheight]{images/figures/dag_bb5.png}

@@

Given the heuristic nature of DAGs, a motivating example can help
illustrate the use of the five fundamental building blocks to construct
a system's causal structure. This example can also help to clarify some
additional conventions for drawing DAGs.

@@

Figure~\ref{fig-example} presents two graphs illustrating the plausible
causal structure of the motivating example.
Figure~\ref{fig-dag_example1} shows the \emph{simplified} DAG. In this
graph, the variables of interest or \emph{endogenous} variables
\(V=\{T,X,Y\}\) are depicted as solid black circles, indicating that
they are observed variables. The arrows in the graph reflect the
expected direction of causal influences among the variables. In
contrast, Figure~\ref{fig-dag_example2} depicts the \emph{magnified}
version of the DAG. This graph includes, in addition to the endogenous
variables, the \emph{exogenous} variables \(E=\{e_{T},e_{X},e_{Y}\}\).
These exogenous variables are commonly referred to as
\emph{disturbances} or \emph{errors}, because they represent factors not
explicitly modeled. They are depicted as open circles, indicating their
unobserved nature. These exogenous variables are often omitted for
simplicity, resulting in the simplified DAG in
Figure~\ref{fig-dag_example1}. However, including them in the graph can
be advantageous in certain scenarios, as they can help to highlight
potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}. These concepts are introduced in the next
section.

@@

Furthermore, a detailed examination of both figures reveals the presence
of at least four of the five fundamental building blocks.
Figure~\ref{fig-dag_example1} displays multiple descendants, evident in
pairwise relations such as \(X \rightarrow T\), \(X \rightarrow Y\), and
\(T \rightarrow Y\). It also illustrates a fork with
\(X \rightarrow \{T,Y\}\). Similarly, Figure~\ref{fig-dag_example2}
features multiple two-unconnected nodes, evident in the pairwise
relations \(\{e_{T} \perp\!\!\!\perp e_{X}\}\),
\(\{e_{T} \perp\!\!\!\perp e_{Y}\}\), and
\(\{e_{X} \perp\!\!\!\perp e_{Y}\}\), as well as colliders such as
\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\). The
symbol \(\perp\!\!\!\perp\) denotes a concept know as
\emph{d-separation}, which roughly implies the independence of the
variables. This concept is introduced in the next section.
